Smart Spreading Factor Assignment for LoRaWANs

### 2018-12-31 paper.tex
proposed teq. conlcusion

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Smart Spreading Factor Assignment for LoRaWAN
}

@@

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over very large geographical areas using license-free frequency bands. Focus on this work is one of the most promising LPWAN technologies: Lora. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum (CSS). LoRa trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Spreading factor (SF) assignment of nodes has strong effect on collisions thus network performance. In this work, we implemented a simulation environment to study different SF assignment schemes and we proposed a machine learning technique to optimize SF assignment. Finally, we illustrate simulation results for proposed machine learning SF assignment technique.
\end{abstract}

@@

\par There are several emerging LPWAN technologies. LoRa, SigFox, NB-IoT and LTE-M are commonly used, well known LPWAN technologies. LoRa and SigFox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost. [CITE] Both LoRa and SigFox known for ultra low power consumption and resilient to interference while NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. [CITE] LoRaWAN and SigFox MAC protocols are based on pure ALOHA medium access. [CITE] LoRaWAN networks can be deployed as private network. However, SigFox and NB-IoT supports only operator contracted deployments. [CITE] Number of messages that Sigfox end device can send in a day is limited to 140 packets for uplink and just 4 packets for downlink. Sigfox packet payload is limited to 12 bytes for uplink and 8 bytes for downlink. [CITE] However LoRa supports up to 243 bytes payload and NB-IoT supports up to 1600 bytes payload. [CITE] Sigfox maximum data rate is also very low 100 bps, on the other hand maximum data rate for LoRa and NB-IoT are 50 kbps and 200 kbps respectively. [CITE]

@@

\par The following two sections provides an background information about LoRaWAN and LoRa. Section \ref{Other Related Works} describes other related works. Section \ref{Proposed Technique} describes proposed technique for SF assignment. Simulation environments and results are shown in Section \ref{Simulation Environment} and \ref{Simulation Results}. Finally, Section \ref{Conclusion} concludes the paper.

@@

\subsection{End node}
\par LoRaWAN end node is low power embedded device that communicates to gateways. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C for providing LPWAN solution to different scenarios. Class A end nodes generates uplink transmission at any time and only receives a period of time after uplink transmission. Class B end nodes extend Class A by adding scheduled receive windows for downlink transmission. Receive window is synchronized using beacon packet transmitted by gateway. Class C end nodes extend Class A by keeping receive window open unless they are transmitting. This provides low latency donwlink communication but requires more power consumption. [CITE] In this paper, we focused on class A end devices since Class A provides lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway is device that receive/transmit packets coming from/to end nodes. A typical gateway can receive from multiple channel at the same time. Gateways are typically connected to power grid so power consumption of a gateway is insignificant most of the deployments.

\subsection{Network server}
\par LoRaWAN network server is server that provides MAC layer processing. Network server routes messages from application to end nodes, and vice versa. Network server can be used for tweaking end node parameters like channel, transmit power, SF to increase network performance.

@@

\par Initially, a node is not aware of how far it is from a GW. However, a node can guess how far away from the gateway by observing received signal power of a downlink transmission. If a downlink transmission received signal power is too high then node can decrease its next transmission SF or transmit power to decrease power consumption.
Also GW can request from a node to decrease node's SF or transmit power. This SD assignment is called lowest possible SF assignemnt scheme for the rest of the paper.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN class A end devices, multiple gateways, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is hugely important for the performance of LoRaWAN networks.

@@

\section{Proposed Technique} \label{Proposed Technique}
\par [TODO]

\par If GW force some of close end nodes to select higher SF even if they can able to communicate with lower SF, this will result lower collisions due to the orthogonality of different SF. Figure~\ref{fig:collision_solution_single_gw}

\par GW can avoid increasing SF of end nodes which is close to other GW using location information obtained by triangulating the signal strength of end nodes. Figure~\ref{fig:collision_solution_multi_gw}

@@

\section{Simulation Environment} \label{Simulation Environment}
\par [TODO] A discrete event simulator is implemented in Python to simulate LoRa network performance. https://github.com/tugrulyatagan/simlorafs

@@

\begin{equation} \label{eq:propagation_loss}
P^{dB}_{PATH} = 120.5 + 37.6 {\log_{10} R|_{km}}
\end{equation}

@@

\begin{table}
\centering
\caption{Gateway Sensitivity for different SFs}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
            &         & \multicolumn{6}{c|}{\textbf{SF}}        \\ \cline{3-8} 
            &         &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
            & 125 kHz & -123 & -126 & -129 & -132 & -133 & -136 \\ 
\textbf{BW} & 250 kHz & -120 & -123 & -125 & -128 & -130 & -133 \\
            & 500 kHz & -116 & -119 & -122 & -125 & -128 & -130 \\
\hline
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{EU863-870 ISM Band LoRa Default Channels}
\label{table:max_tx_power}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Channel} & \textbf{Frequency} & \textbf{Bandwidth} & \textbf{Max EIRP} \\ \hline
      1 &     868.1 &   125 kHz &   14 dBm \\
      2 &     868.3 &   125 kHz &   14 dBm \\
      3 &     868.5 &   125 kHz &   14 dBm \\

@@

\subsection{Simulation Assumptions}
\par [TODO]

@@

\section{Simulation Results} \label{Simulation Results}
\par [TODO] Evaluation

\section{Conclusion} \label{Conclusion}
\par [TODO] Summary, future works
\cite{7815384} \cite{7803607} \cite{7996384} \cite{8090518} \cite{s17061193} \cite{8267219} \cite{8430542} \cite{8319183} \cite{8480649} \cite{AN1200.22} \cite{Bor:2016:LLW:2988287.2989163} \cite{8406255} \cite{DBLP:journals/corr/abs-1802-10338} \cite{finnegan2018comparative}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{multirow}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Smart Spreading Factor Assignment for LoRaWANs}

@@

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over very large geographical areas using license-free frequency bands. Focus on this work is one of the most promising LPWAN technologies: LoRa. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum. LoRa trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Spreading factor (SF) assignment of nodes has strong effect on collisions thus network performance. In this work, we implemented a simulation environment to study different SF assignment schemes and we proposed a machine learning technique to optimize SF assignment. Finally, we illustrate simulation results for proposed machine learning SF assignment technique.
\end{abstract}

@@

\par There are several emerging LPWAN technologies. LoRa, Sigfox, NB-IoT and LTE-M are commonly used, well known LPWAN technologies. LoRa and Sigfox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost. [CITE] Both LoRa and Sigfox known for ultra low power consumption and resilient to interference while NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. [CITE] LoRaWAN and Sigfox MAC protocols are based on pure ALOHA medium access. [CITE] LoRaWAN networks can be deployed as private network. However, Sigfox and NB-IoT supports only operator contracted deployments. [CITE] Number of messages that Sigfox end device can send in a day is limited to 140 packets for uplink and just 4 packets for downlink. Sigfox packet payload is limited to 12 bytes for uplink and 8 bytes for downlink. [CITE] However LoRa supports up to 243 bytes payload and NB-IoT supports up to 1600 bytes payload. [CITE] Sigfox maximum data rate is also very low 100 bps, on the other hand maximum data rate for LoRa and NB-IoT are 50 kbps and 200 kbps respectively. [CITE]

@@

\par The following two sections provides an background information about LoRaWAN and LoRa. Section \ref{Other Related Works} describes other related works. Section \ref{Proposed Technique} describes proposed technique for SF assignment. Simulation environments and results are shown in Section \ref{Simulation Environment and Results}. Finally, Section \ref{Conclusion} concludes the paper.

@@

\subsection{End node}
\par LoRaWAN end node (EN) is low power embedded device that communicates to gateways. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C for providing LPWAN solution to different scenarios. Class A end nodes generates uplink transmission at any time and only receives a period of time after uplink transmission. Class B end nodes extend Class A by adding scheduled receive windows for downlink transmission. Receive window is synchronized using beacon packet transmitted by gateway. Class C end nodes extend Class A by keeping receive window open unless they are transmitting. This provides low latency donwlink communication but requires more power consumption. [CITE] In this paper, we focused on Class A end devices since Class A provides lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is device that receive/transmit packets coming from/to end nodes. A typical gateway can receive from multiple channel at the same time. Gateways are typically connected to power grid so power consumption of a gateway is insignificant most of the deployments.

\subsection{Network server}
\par LoRaWAN network server (NS) is server that provides MAC layer processing. Network server routes messages from application to end nodes, and vice versa. Network server can be used for tweaking end node parameters like channel, transmit power, SF to increase network performance.

@@

\par Initially, a node is not aware of how far it is from a GW. However, a node can guess how far away from the gateway by observing received signal power of a downlink transmission. If a downlink transmission received signal power is too high then node can decrease its next transmission SF or transmit power to decrease power consumption. Also GW can request from a node to decrease node's SF or transmit power. This SF assignment method is called lowest possible SF assignment scheme for the rest of the paper.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple gateways, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is hugely important for the performance of LoRaWAN networks.

@@

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure~\ref{fig:collision} can be solved if the GW force some of the close end nodes to select higher SF even if they can able to communicate with lower SF. This may result lower collisions due to the orthogonality of different SF as shown in Figure~\ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure~\ref{fig:collision_solution_single_gw}. However, how GW assign SF to which node becomes an another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collision number with nodes in other GW's range. Thus, extra care must be taken for nodes in intersection area of GWs illustrated in Figure~\ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collision number.  This technique learns nodes behavior of a network. A NS can keep track of successful uplink transmission from nodes and their SFs. NS can also keep track of some of the collided transmission if header part of the packet does not interfered at GW. However, NS cannot keep track of transmissions with receive power lower than GW sensitivity. Using these obtained information NS can train a classifier for predicting transmission result for specific node and specific SF input. Using this predictor GWs can assign the lowest possible SF to nodes considering the collision probability.

\par In this work, we used decision tree classifier to predict transmission results. We utilized Gini index criteria to measure quality of decision tree splits and used balanced class weights. As most of the machine learning applications, increasing data set size increases prediction accuracy. For our case, data set size is directly proportional to simulation duration. Thus increasing simulation duration, increases the prediction accuracy. For real world cases, NS can keep track of transmission and it can create a classifier daily basis, then GW can request from nodes to use this new SF.

@@

\section{Simulation Environment and Results} \label{Simulation Environment and Results}
\par A discrete event simulator is implemented in Python to study LoRa network performance. The LoRaWAN SF simulation tool is publicly available at https://github.com/tugrulyatagan/simlorafs.

Simulation Assumptions

@@

\begin{equation} \label{eq:propagation_loss_simplified}
P^{dB}_{PATH} = 120.5 + 37.6 {\log_{10} R|_{km}}
\end{equation}

@@

\begin{table}
\centering
\caption{Gateway Sensitivity for different SFs}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW}}  & 125 kHz & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                              & 250 kHz & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                              & 500 kHz & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{EU863-870 ISM Band LoRa Default Channels}
\label{table:max_tx_power}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Channel} & \textbf{Frequency} & \textbf{Bandwidth} & \textbf{Max EIRP} \\ \hline
      1 &     868.1 &   125 kHz &   14 dBm \\ \hline
      2 &     868.3 &   125 kHz &   14 dBm \\ \hline
      3 &     868.5 &   125 kHz &   14 dBm \\

@@

@@

\subsection{Simulation Results}
\par [TODO] Evaluation

\section{Conclusion} \label{Conclusion}
\par In this paper, after brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance up to [TODO] \%.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume nodes always uses maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmission more vulnerable to interference thus requires extra care. Optimizing another node parameter alongside SF can be interesting research. Also other machine learning methods can be investigated for SF optimization. 
Reinforcement learning can be a good candidate for this network optimization issue.

\par \cite{7815384} \cite{7803607} \cite{7996384} \cite{8090518} \cite{s17061193} \cite{8267219} \cite{8430542} \cite{8319183} \cite{8480649} \cite{AN1200.22} \cite{Bor:2016:LLW:2988287.2989163} \cite{8406255} \cite{DBLP:journals/corr/abs-1802-10338} \cite{finnegan2018comparative}
